---
title: "Machine Learning Course Project"
author: "Patricio Villar"
date: "Tuesday, November 18, 2014"
output: html_document
---

## Synopsis

This study focuses on weight lifting exercises simulation. Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience. We made sure that all participants could easily simulate the mistakes in a safe and controlled manner by using a relatively light dumbbell (1.25kg)

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz3Jv3tWGMh

One dataset was provided for training and testing purposes, and one separate 20 cases dataset has been provided by Jeff Leek to predict exercise execution types.

## Data Acquisition and Cleaning:

```{r}
## load required lybraries
library(caret)
library(RCurl)
library(randomForest)
 
#Acquire data from public repos:
trainURL <- getURL("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", ssl.verifypeer=0L, followlocation=1L)
testURL <- getURL("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", ssl.verifypeer=0L, followlocation=1L)
writeLines(trainURL, "training.csv")
writeLines(testURL, "testing.csv")
training <- read.csv("training.csv", header=TRUE, na.strings=c("NA", ""))
testing <- read.csv("testing.csv", header=TRUE,na.strings=c("NA", ""))
```

Then, we proceed to clean data and select features, by getting rid of NA columns and filtering the non-descriptive first seven columns:


```{r}

#PreProcess Training data - eliminate NA columns and first 7 columns (irrelevant data)

na.index <- which(is.na(training[1,]), arr.ind = T)
clean_training <- training[,-na.index]
clean_training <- clean_training[8:59]
clean_testing <- testing[,-na.index]
clean_testing <- clean_testing[8:59]
```

Then, we verify data the contains little to no variance, looking for covariates to rule out:

```{r, echo=TRUE}

NZV <- nearZeroVar(x = clean_training, saveMetrics = TRUE)
print(NZV)
```

Since all the variables provide considerable variability, we decide to leave them all in place.

## Data Partitioning

Since we've been provided with a large dataset (~19K observations), we decided to split it in 2 smaller datasets, each one with his own training and testing section (with a standard training/testing ratio of 60/40):


```{r, echo=TRUE}

##divide in 2 data sets and each ds divided by train and test dss:
set.seed(1234)
part1 <- createDataPartition(clean_training$classe, p=0.5, list=FALSE)
ds1 <- clean_training[part1,]
ds2 <- clean_training[-part1,]

set.seed(1234)
inTrain <- createDataPartition(ds1$classe, p=0.6, list=FALSE)
training1 <- ds1[inTrain,]
testing1 <- ds1[-inTrain,]

set.seed(1234)
inTrain <- createDataPartition(ds2$classe, p=0.6, list=FALSE)
training2 <- ds2[inTrain,]
testing2 <- ds2[-inTrain,]

```

## Algorithm:

We decided to use Random Forest with cross validation

```{r, echo=TRUE}

## Apply Random Forest with cross validation

set.seed(1234)

modelRF <- train(training1$classe ~ ., method = "rf", trControl=trainControl(method="cv", number=2), data=training1)
print(modelRF)
```

Since we've got an accuracy level of 0.98, which is pretty good we decided to move forward and test it with our first testing set:

```{r, echo=TRUE}
pred1 <- predict(modelRF, newdata=testing1)
print(confusionMatrix(pred1, testing1$classe), digits=2)
```

Which gave us a good result, with a very low "out of sample error":

Out of sample error (Random Forest + Cross Validation) = 1 - 0.98 = 0.02

## Evaluation:

After building our machine learning block, we got to apply it to our testing set so we can predict outcomes and test it against J. Leek data set:

```{r, echo=TRUE}
## test with the real testing set

pred2 <- predict(modelRF, newdata=clean_testing)
print(pred2)
```

Which gave us a pretty good result, with 19/20 accurate responses.
